\documentclass{article}

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}

\title{Probability Theory Homework 5}
\author{Gregory Matsnev}
\date{\today}

\begin{document}

\maketitle

\section{Problem 1}

\textbf{Problem.} Find the PDF of $Y = 1 - X^3$, where $X$ is the random variable distributed according to the Cauchy law, i.\ e.\ with the PDF
\[
\phi(x) = \frac{1}{\pi(1 + x^2)}
\]

\textbf{Solution.}
Let
\[
Y=1-X^3.
\]
Then
\[
y=1-x^3 \quad \Longleftrightarrow \quad x^3=1-y \quad \Longleftrightarrow \quad x=(1-y)^{1/3}.
\]
Computing the CDF:
\[
F_Y(y)=\mathbb P(Y\le y)=\mathbb P(1-X^3\le y)
=\mathbb P(X^3\ge 1-y)=\mathbb P\!\left(X\ge (1-y)^{1/3}\right)
=1-F_X\!\left((1-y)^{1/3}\right).
\]
Differentiating:
\[
f_Y(y)=\frac{d}{dy}F_Y(y)
=-\frac{d}{dy}F_X\!\left((1-y)^{1/3}\right)
=-f_X\!\left((1-y)^{1/3}\right)\frac{d}{dy}(1-y)^{1/3}.
\]
\[
\frac{d}{dy}(1-y)^{1/3}=-\frac{1}{3}(1-y)^{-2/3},
\qquad
\left|\frac{d}{dy}(1-y)^{1/3}\right|=\frac{1}{3|1-y|^{2/3}}.
\]
Hence
\[
f_Y(y)=f_X\!\left((1-y)^{1/3}\right)\frac{1}{3|1-y|^{2/3}}.
\]
Substituting $f_X$:
\[
f_X\!\left((1-y)^{1/3}\right)=\frac{1}{\pi\left(1+\left((1-y)^{1/3}\right)^2\right)}
=\frac{1}{\pi\left(1+|1-y|^{2/3}\right)}.
\]
Therefore
\[
f_Y(y)=\frac{1}{\pi\left(1+|1-y|^{2/3}\right)}\cdot \frac{1}{3|1-y|^{2/3}}
=\boxed{\ \frac{1}{3\pi\,|1-y|^{2/3}\left(1+|1-y|^{2/3}\right)}\ },\qquad y\in\mathbb R.
\]

\section{Problem 2}

\textbf{Problem.} Find the expected value and the variance of the random variable $Y = 2 - 3\sin X$, given that the PDF of $X$ is
\[
\phi(x) = \frac{1}{2}\cos x \text{ for } x \in \left[-\pi/2,\pi/2\right]
\]

\textbf{Solution.}

Letting $f_X(x)=\frac12\cos x$ for $x\in\left[-\frac\pi2,\frac\pi2\right]$.

Computing:
\[
\mathbb E[Y]=\mathbb E[2-3\sin X]
=\int_{-\pi/2}^{\pi/2}(2-3\sin x)f_X(x)\,dx
=\int_{-\pi/2}^{\pi/2}(2-3\sin x)\frac12\cos x\,dx
\]
\[
=\int_{-\pi/2}^{\pi/2}\cos x\,dx-\frac32\int_{-\pi/2}^{\pi/2}\sin x\cos x\,dx
=\Bigl[\sin x\Bigr]_{-\pi/2}^{\pi/2}-\frac32\Bigl[\frac12\sin^2 x\Bigr]_{-\pi/2}^{\pi/2}
=2.
\]

Computing:
\[
\mathbb E[Y^2]=\mathbb E[(2-3\sin X)^2]
=\int_{-\pi/2}^{\pi/2}(2-3\sin x)^2 f_X(x)\,dx
=\int_{-\pi/2}^{\pi/2}(4-12\sin x+9\sin^2 x)\frac12\cos x\,dx
\]
\[
=2\int_{-\pi/2}^{\pi/2}\cos x\,dx-6\int_{-\pi/2}^{\pi/2}\sin x\cos x\,dx+\frac92\int_{-\pi/2}^{\pi/2}\sin^2 x\cos x\,dx.
\]
\[
2\int_{-\pi/2}^{\pi/2}\cos x\,dx=2\Bigl[\sin x\Bigr]_{-\pi/2}^{\pi/2}=4,
\qquad
-6\int_{-\pi/2}^{\pi/2}\sin x\cos x\,dx=-6\Bigl[\frac12\sin^2 x\Bigr]_{-\pi/2}^{\pi/2}=0.
\]
\[
\frac92\int_{-\pi/2}^{\pi/2}\sin^2 x\cos x\,dx
=\frac92\int_{-1}^{1}u^2\,du
=\frac92\Bigl[\frac{u^3}{3}\Bigr]_{-1}^{1}
=3.
\]
\[
\mathbb E[Y^2]=4+0+3=7,
\qquad
\mathrm{Var}(Y)=\mathbb E[Y^2]-\bigl(\mathbb E[Y]\bigr)^2=7-2^2=3.
\]

\section{Problem 3}

\textbf{Problem.} The random variable $X$ is defined on the entire real axis with the probability density $\phi(x)=\frac{1}{2}e^{-|x|}$. Find the probability density of the random variable $Y=X^2$ and its mathematical expectation.

\textbf{Solution.}

\section{Problem 4}

\textbf{Problem.} Prove formally that if the correlation coefficient $\rho_{XY}$ of two random variables $X$ and $Y$ is equal in absolute value to one, then there is a linear functional relationship between these random variables.

Remember how to prove that $Cov(X,Y)\leq \sigma_X\sigma_Y$.

\textbf{Solution.}

\section{Problem 5}

\textbf{Problem.} The distribution surface (joint PDF) of the two-dimensional random variable $(X,Y)$ is a right circular cone, the base of which is a circle centered at the origin with a unit radius. Outside this circle, the joint PDF of this two-dimensional random variable $(X,Y)$ is zero. Find the joint PDF $f(x,y)$, the marginal PDFs and the conditional PDFs $f_x(y)$ and $f_y(x)$. Are the random variables $X$ and $Y$ dependent and/or correlated?

\textbf{Solution.}

\section{Problem 6}

\textbf{Problem.} Let $X$ and $Y$ be continuous random variables with a (spherically symmetric) joint PDF of the form $f(x,y)=g(x^2+y^2)$ for some function $g$. Let $(R,\theta)$ be the polar coordinates of $(X,Y)$, so that $R^2=X^2+Y^2$ is the squared distance from the origin and $\theta$ is the angle $\in[0,2\pi)$, with $X=R\cos\theta$, $Y=R\sin\theta$.

a) Prove that $R$ and $\theta$ are independent and explain intuitively why this result makes sense;

b) What is the joint PDF of $(R,\theta)$ if $(X,Y)$ is Uniform on the unit disk, i.\ e.\ $x^2+y^2\leq 1$? If $X,Y$ are i.\ i.\ d.\ $N(0,1)$?

\textbf{Solution.}

\end{document}
